---
title: "Fall_2021_R_workshopWHOI"
author: "Anya Brown"
date: "10/6/2021"
output: html_document
---

```{r setup, include=FALSE}
coat <- c("calico","black","tabby")
weight <- as.numeric(c(2.1, 5.0, 3.2))
like_string <- c(1,0,1)

cats <- cbind(coat, weight, like_string)
cats <- as.data.frame(cats)
```


```{r}
cats$coat <- as.factor(cats$coat)
cats$weight <- as.numeric(as.character(cats$weight))
cats$like_string <- as.integer(cats$like_string)
cats$coat_car <- as.character(cats$coat)
#convert column to character to change something
cats$coat_car[2] <- "orange"
```

```{r}
new_column <- c(1,2,3)
cats_new <- cbind(cats, new_column)
```

```{r}
cats_new_bigger <- rbind(cats_new, list("tabby",3.2, 1,3))
```


Subsetting data


```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')

```



```{r}

x <- c(a =5.4,b=6.2, c = 7.1, d = 4.8, e = 7.5)
```

Subset vector
```{r}

x1 <- x[1]
xb <- x["b"]

```
Conditional subsetting

Logical vector
```{r}
logicalsubset <- x[c(TRUE,FALSE, FALSE,TRUE,FALSE)]
```
subset based on values
```{r}
x == 6.2
x[x==6.2]
x[x<10]
x[x<6]
```
Combining conditionals
& = needs to be true for both conditions
```{r}

x[x<6 & x >5]

```
OR operator: |
one thing needs to be true
```{r}
x[x<6 | x>5]

```

Challenge 2
values in vector greater than 4, less than 7
```{r}
x[x>4 & x<7]
```

Challenge 3 - using the gapminder data
```{r}
gapminder <- read.csv("data/gapminder_data.csv", header = T)
seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
```

```{r}
countries <- unique(gapminder$country)
SEAsiacountries <- countries[countries == "Myanmar" | countries == "Thailand" | countries == "Cambodia" | countries == "Vietnam" | countries == "Laos"]
```

Usng %in% !!!
```{r}
countries %in%  seAsia

SEgapminder <- gapminder[which(gapminder$country %in% seAsia),]

```
Practicing if then statements
```{r}
#if (exp){
  do this if TRUE
#}
#else{
#  do this if false
#}
  
```

Conditional statements in R - uses a single true/false...use ifelse if there are multiple rows
```{r}
x <- 3

if(x > 10){
"Yay!"  
} else if (x > 5 & x < 10){
  "Boo"
} else {
  "Low number"
}

```
gapminder records from 2002
```{r}
nrow2002 <- nrow(gapminder[gapminder$year == 2002,])
#test <- gapminder[gapminder$year == 2002,]
test <- nrow2002 > 0
if(test){
  "2002!"
}
```
alternate using all and any
all = if everything in the result is one way, return a value
any = if anything is true return that value
```{r}

test2 <- any(gapminder$year == 2002)

if(test2){
  "2002!"
}

```
Iteration: FOR LOOPS!!!!
```{r}
#for(element in a set of values){
 # do the stuff
#}

```

example
```{r}
for(i in 1:10){
  print(i)
}
```
Example: set up
```{r}
x <- 1:10
x+1
```
Predefine output at the size it should be at teh end - helps with computing time, important for large datasets
```{r}
out <- integer(length(x)) # could also include length = length(x)
for(i in x){
  out[i] <- i + 1
}
out
```
While loops - not commonly used in R 
```{r}
while(true){
  do this
}
```

```{r}
library(dplyr)
gapminder = read.csv("data/gapminder_data.csv")

```
```{r base_r}
#pull out rows and a certain column
mean(gapminder[gapminder$continent == "Africa","gdpPercap"])

```
```{r}
mean(gapminder[gapminder$continent == "Asia","gdpPercap"])
```
Using Select
```{r}

year_country_gdp <- dplyr::select(gapminder, year, country, gdpPercap)
```

```{r}
dplyr::select(gapminder, -continent)
```
```{r using pipe}

year_country_gdp <- gapminder %>% select(year, country, gdpPercap)

```
Rename old column name to new name
```{r rename}
year_country_gdp %>% rename(gdp_per_capita = gdpPercap)
```
```{r filter}

gapminder %>% filter(continent == "Europe", year == 2007) %>% select(year, country, gdpPercap)
  

```
using group to chunk the data
```{r group_by}
str(gapminder)
```

```{r group_by2}
str(gapminder %>% group_by(continent))
```
```{r}
gapminder %>% group_by(continent, year)%>%
  summarize(mean_gdp_per_capita = mean(gdpPercap), sd_gdp_per_capita = sd(gdpPercap))
  
```

```{r}
gapminder %>% 
  filter(year == 2002) %>%
  count(continent, sort = TRUE)
```

```{r se}
gapminder %>% group_by(continent) %>%
  summarize(se = sd(lifeExp)/sqrt(n()))
```

```{r if else}
gdp_pop_bycontinents_byyear <- gapminder %>%
    mutate(gdp_billion = gdpPercap*pop/10^9) %>%
    group_by(continent,year) %>%
    summarize(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop),
              mean_gdp_billion = mean(gdp_billion),
              sd_gdp_billion = sd(gdp_billion))


# calculate GDP only for people with a life expectation above 25
gdp_pop_bycontinents_byyear_above25 <- gapminder %>%
    mutate(gdp_billion = ifelse(lifeExp > 25, gdpPercap * pop / 10^9, NA)) %>%
    group_by(continent, year) %>%
    summarize(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop),
              mean_gdp_billion = mean(gdp_billion),
              sd_gdp_billion = sd(gdp_billion))

```

```{r}
library(ggplot2)
```

```{r}
afgan_data <- gapminder %>% 
  filter(country == "Afganistan")
```
```{r plot data}
ggplot(afgan_data, mapping= aes(x = year, y = lifeExp)) + geom_line()

```
```{r}
asia_data <- gapminder %>% filter(continent == "Asia")
```
```{r}
ggplot(asia_data, aes(x = year, y = lifeExp)) + geom_line(aes(color = country))
```

FUNCTIONS!
Why and when do we want to write a function?
if you copy and paste code more than two times

name <- function(parameters){
do some stuff
return(object)
}
```{r}
fahr_to_kel <- function(temp){
  kel <- ((temp-32)*5/9) + 273.15
return(kel)
}

fahr_to_kel(212)
fahr_to_kel(32)
```

```{r K to C}
kel_to_cel <- function(temp) {
  cel <- temp-273.15
  cel
}

```

```{r}
kel_to_cel(273)
```

```{r nested functions}
fahr_to_cel <- function(temp) {
  if(!is.numeric(temp)){
    stop("temp must be a numeric vector")
  }
  kel <- fahr_to_kel(temp)
  cel <- kel_to_cel(kel)
  cel
}


fahr_to_cel(32)
fahr_to_cel("blue")
fahr_to_cel(212)
```
```{r}
calcGDP <- function(pop, gdpPercap) {
  pop * gdpPercap
 
}
```

```{r}
calcGDP(1,1)
```

```{r}
#shift, control, m
gapminder %>% 
  mutate(gdp = calcGDP(pop, gdpPercap))
```

